All of the topics covered in this course make extensive use of molecular symmetry and group theory which are, by themselves, topics of entire books.
It is the aim of this text to present group theory in sufficient detail so that the student can solve most day-to-day spectroscopic problems and read the literature with confidence.

A \keyword{symmetry operation} is a rotation and/or reflection which leaves the molecule unchanged. 
It is performed about a \keyword{symmetry element}---a point, a line or a plane.
The square planar \ce{PtCl4^{2-}} ion is reported to be a highly symmetric ion because it contains a large number of symmetry elements.
All of the symmetry operations that apply to a molecule or ion constitute its \textit{point group}.
It is the point group to which the molecule belongs that designates its symmetry.
The following section will exhibit sixteen symmetry operations that can be performed on the \ce{PtCl4^{2-}} ion. 
These operations constitute what is known as the \PG{D}{4h} point group, i.e. \ce{PtCl4^{2-}} has \PG{D}{4h} symmetry.
In a treatment of the spectroscopy or bondings of an inorganic compound, one must first determine the point group to which the system belongs.
A full appreciation of what \PG{D}{4h} symmetry implies requires the visualisation of the symmetry elements and operations as well as an understanding of groups.
In the following two sections thereafter, various symmetry operations as well as groups will be discussed.
The remainder of the chapter is devoted to representations of the point groups and to some simple applications of group theory to spectroscopy and bonding.

\input{chapters/01-symmetry/01-symops-and-elements.tex}
\input{chapters/01-symmetry/02-groups.tex}
\input{chapters/01-symmetry/03-similarity-transformations.tex}
\input{chapters/01-symmetry/04-point-groups.tex}
% \input{chapters/01-symmetry/05-matrix-representation.tex}
